---
title: "Import Parras-Molto2018 mock abundance table into phyloseq"
output:
  html_document:
    toc: false
    toc_float: false
    self_contained: true
    code_folding: show
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_knit$set(progress=TRUE, verbose=TRUE)
# Global chunk options
knitr::opts_chunk$set(cache=FALSE, echo=TRUE, warning=FALSE,
    message=FALSE, include=TRUE)
# R options
# options(width=100)
options(stringsAsFactors = FALSE)
```

Goal is to create a `phyloseq` object with the viral abundance table that was
used to create Figure 1.

# Setup

Load packages.
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)
```
Set paths.
```{r path}
dotenv::load_dot_env("../.env")
script_path <- getwd()
data_path <- file.path(Sys.getenv("DATA_DIR"),
    "parras-molto2018")
```

# Download supplemental files

[Article](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0507-3)

Additional file 1: Table S1. Overview of samples and procedures followed for
the analysis of mock communities and saliva samples. (XLSX 11 kb)
https://static-content.springer.com/esm/art%3A10.1186%2Fs40168-018-0507-3/MediaObjects/40168_2018_507_MOESM1_ESM.xlsx

"Additional file 4: Table S4. Effect of several treatments in bacteria removal.
(XLSX 9 kb)"
https://static-content.springer.com/esm/art%3A10.1186%2Fs40168-018-0507-3/MediaObjects/40168_2018_507_MOESM4_ESM.xlsx

"Additional file 5: Table S5. Virus enrichment and random DNA amplification
effects on mock viral communities. Number of genomes determined by absolute
qPCR. (XLSX 14 kb)"
https://static-content.springer.com/esm/art%3A10.1186%2Fs40168-018-0507-3/MediaObjects/40168_2018_507_MOESM5_ESM.xlsx


```{r}
file_num <- c(1, 4, 5)
urls <- paste0("https://static-content.springer.com/esm/",
    "art%3A10.1186%2Fs40168-018-0507-3/MediaObjects/",
    "40168_2018_507_MOESM", file_num, "_ESM.xlsx")
fns <- file.path(data_path, "supplementary_files", 
    paste0("additional_file_", file_num, ".xlsx"))
names(fns) <- file_num
dir.create(file.path(data_path, "supplementary_files"), recursive = TRUE)
walk2(urls, fns, download.file)
```

# Create phyloseq for viruses

First, we need to import the abundance data from the Excel file. The table for
the first mock community is in range "A2:H21", including a header row with the
taxa names. The second row is not data, but only has the label "Mock community
1" and so can be removed by filtering to rows with non-NA abundances. The table
for the second mock (without header) and the row labeled "Mock community 2" is
in range "A23:H47".
```{r}
mock1 <- readxl::read_xlsx(fns["5"], range = "A2:H21") %>%
    rename(Sample = X__1) %>%
    filter(!is.na(Lambda))
mock2 <- readxl::read_xlsx(fns["5"], range = "A23:H47", 
    col_names = colnames(mock1))
```
To give each sample unique names in the combined table, I will prepend "M1" and
"M2" to the sample names.
```{r}
mock_tb <- list("1" = mock1, "2" = mock2) %>%
    bind_rows(.id = "Mock") %>%
    mutate(Sample = paste0("M", Mock, "_", Sample))
```

Next, we want to parse sample metadata from the sample names and mock community
number into a `sample_data` dataframe.
The sample name has the format "Treatment_#" where "#" is the replicate number.
```{r}
sam_tb <- mock_tb$Sample %>%
    str_match("M([1-2])_(.+)_([1-9])") %>%
    as_tibble() %>%
    rename(Sample = V1, Mock = V2, Treatment = V3, Replicate = V4)
```

Next, we could optionally parse the treatment strings to get a column for each
atomic treatment.
```{r}
```

Then we make a phyloseq object:
```{r}
taxa <- colnames(mock1)[-1]
print(taxa)
abun <- mock_tb %>%
    select(taxa) %>%
    data.frame(row.names = mock_tb$Sample) %>%
    otu_table(taxa_are_rows = FALSE)
head(abun)
sam <- sam_tb %>%
    select(-Sample) %>%
    data.frame(row.names = sam_tb$Sample) %>%
    sample_data
head(sam)
ps <- phyloseq(abun, sam)
dir.create(file.path(data_path, "final"))
saveRDS(ps, file.path(data_path, "final", "phyloseq.Rds"))
```

